Stemmer Tool Manual

### 2018-03-14 tex/merf/OldSubmission/Apendix.tex
old submission files

\begin{figure*}[tb]
  \centering
  \includegraphics[width=0.9\textwidth]{figures/mbfedit}
  \caption{\framework tag type Boolean formula editor.}
  \label{f:bfe}
\end{figure*}

\framework provides a user friendly interface to specify the 
atomic terms, the \framework Boolean formulae, the \framework regular expressions, 
the tag types, and the legends. 
The \framework GUI also allows the user to modify and correct the 
resulting tag set $R$.
The \framework GUI allows the user also to compute accuracy results 
that compare different tag sets. 
The accuracy results serve well as inter annotation agreement results
when the tag sets come from two human annotators, 
or as evaluation results for learning and information extraction techniques.

The snapshot in Figure~\ref{f:tagger} shows the \framework GUI with 
the tag type color sensitive view, the tag list view, the tag description view, 
and the match tree. 
In the shown example, the user specified three MBF-based tag types with labels ``HUNDRED'',
``KEY'', and ``DIGITSTENS''. 
The ``HUNDRED'' tag type is based on a simple MBF formula that inspects the gloss tag of 
the stem of the word and checks whether it is {\tt hundred}. 
The ``KEY'' tag type is built on an MBF formula that computes a disjunction 
between the gloss tags of the stems being {\tt thousand/s}, {\tt million/s}, and {\tt billion/s}. 
The ``DIGITSTENS'' tag type is based on a disjunction formula between the gloss tags of the stems 
being {\tt one}, {\tt two},\dots, {\tt ten}, and {\tt twenty}, {\tt thirty},\dots, {\tt ninety}. 
The user also defined an MRE-based tag type with the label ``number''. 
The ``number'' tag type is based on the formula $\left(\textit{HUNDRED}|\textit{KEY}|\textit{DIGITSTENS}\right)+$. 
In words, the formula defines the occurence of one or more consequent tags of the three tags.

The context sensitive menus in Figure~\ref{f:tagger} allow the user to tag 
a selected word or chunk differently or to entirely remove the tag.

The \framework GUI also allows manual tag types that are not based on morphological features.
These tags enable the users to build their own reference corpora without 
help from the morphological analyzer.

\begin{figure*}[tb]
  \centering
  \includegraphics[width=0.9\textwidth]{figures/msfedit}
  \caption{\framework tag type regular expression editor.}
  \label{f:sfe}
\end{figure*}

\subsection{Tag type Boolean formula editor}
The morphology based tag type editor shown in Figure~\ref{f:bfe}
allows the user to write a tag type Boolean formula in a user friendly manner. 
The user first specifies atomic terms by selecting 
a feature from ${\cal F}$. 
The pattern is a regular expression that filters the feature values.
This implicitly helps the user specify 
an {\em exact} versus a {\em partial} match of the  desired 
morphological feature value.

Then the user can add the selected feature values to the atomic terms 
under the tag type name. 
The feature column has a context sensitive menu that allows negating 
the term.
The value column has a context sensitive menu that can switch the 
operation between the values in 
${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$.
The right pane shows a description of the tag type and a set of legend 
descriptors.

For example, the user can select the feature {\em category} and associate it 
with the value {\em temporal unit}. 
Multiple feature/value pairs can be included in a single tag type 
definition with a disjunction semantics.

\subsection{Tag type regular expression editor}
The morphology based tag type editor shown in Figure~\ref{f:sfe} allows the 
user to define a tag type regular expression in a user friendly manner. 
The user first adds the required MREs to the formula under the tag type name 
by selecting a label from ${\cal T}$ under MBFs. 
The Boolean formula of a highlighted tag type is shown in the table on the lower left pane. 
Each selected label is associated with an automatic name. 
The regular expression tree features the name, label, and operation for each member MRE.

{\tt NONE} is a special tag type with no Boolean formula. 
It tags all the words that are not tagged with any of the morphology based tag types. 
{\tt NONE} is defined by default and can be used to introduce flexibility and noise tolerance into the formula. 
Moreover, a defined MRE is directly added to the MRE list enabling recursion 
in formula definition.

Then, the user can construct the formula by applying operations to the selected MREs. 
To do so, the user selects one or two MREs then selects an operation from the ones shown 
in the lower left of the view. 
Without adding operations, the formula is a sequence of the selected MREs. 
The right pane shows a description of the tag type and a set of legend 
descriptors.

For example, the user can add the {\tt DIGITSTENS}, {\tt KEY}, and {\tt HUNDRED} MREs to the formula. 
Then, he/she selects the {\tt DIGITSTENS} and {\tt KEY} MREs and apply the disjunction ( $|$ ) 
operation to them. 
Similarly, he/she can apply the disjunction operation to the disjunction MRE formed before and {\tt HUNDRED}. 
Last, the user selects the disjunction formula and applies the Kleene star. 
Thus, we get the regular expression shown in Figure~\ref{f:sfe}.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.45\textwidth]{figures/actions}
  \caption{\framework MRE action sample.}
  \label{f:actions}
\end{figure}

In order to add computational actions to an MRE, 
the user selects to edit the actions in the MRE tag type editor. 
The view shown in Figure~\ref{f:actions} allows the user to specify 
the actions in a user friendly manner. 
\framework enables the user to access the MBF match features easily and define the pre-match actions. 
\framework also enables the user to edit the formula declarations and library includes.

\subsection{Analysis}
In addition to automatic and manual tagging, \framework 
allows comparing tag sets and tag types applied to the
same input text. 
The \framework comparator takes as input two tag sets $R_1$ and $R_2$ and 
two tag type sets ${\cal T}_1$ and ${\cal T}_2$. 
It produces a difference view for the tag types and 
a difference view for the tag sets. 
The tag type difference view shows the common tag types ${\cal T}_1 \cap {\cal T}_2$,
the tag types in ${\cal T}_1$ and not in ${\cal T}_2$, 
and the tag types in ${\cal T}_2$ and not in ${\cal T}_1$.

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{figures/diff}
  \caption{\framework comparison and accuracy results view.}
  \label{f:diff}
\end{figure*}

@@

>>>>

\subsection{Finite state transducers}

A {\em finite state transducer} (FST) is a finite state machine with input and output tapes. 
FSTs differ from {\em finite state automata} (FSA) in that
they have an output tape while FSAs have accept states instead.

Formally, an FST is a tuple $M = (S,S_{0},\sigma,\varGamma,\delta)$ where $S$ is the set of states,
$S_{0} \subset S$ is the set of initial states, 
$\sigma$ is the input alphabet, 
$\varGamma$ is the output alphabet, 
and $\delta \subseteq S \times (\sigma \times \{\epsilon\}) \times (\varGamma \times \{\epsilon\}) \times S$ is the transition relation.

FSTs have been used extensively in text mining applications where the input is the text and the output is the
delimiters of a chunk of text with an associated class~\cite{beesley2001finite}. 
FSTs are attractive in NLP tasks due to their efficiency and ease of use.

\subsection{Classes, labels, and tag types}

A {\em class} is a semantic decision that the NLP of CL task tries to make. 
Parts of text that belong to the desired class are all assigned the same class {\em label}. 
For example, {\em temporal unit} is a label of a class that encompasses explicit temporal 
information in the sentence related to time such as 
\RL{dqyqT} (minute) and \RL{sA`T} (hour).

The user may wish to define a simple class as an abstract category that contains a 
set of morphemes. 
For more sophisticated classes, the user can use the visual interface in \framework to 
define the class through Boolean formulae with morphology-based atomic terms. 
The annotation of text with the labels of the classes can later be used in 
CL and NLP tasks for learning, testing, and validation.

\subsection{Analysis}

In addition to automatic and manual tagging, \framework 
allows comparing tag sets and tag types applied to the
same input text. 
The \framework comparator takes as input two tag sets $R_1$ and $R_2$ and 
two tag type sets ${\cal T}_1$ and ${\cal T}_2$. 
It produces a difference view for the tag types and 
a difference view for the tag sets. 
The tag type difference view shows the common tag types ${\cal T}_1 \cap {\cal T}_2$,
the tag types in ${\cal T}_1$ and not in ${\cal T}_2$, 
and the tag types in ${\cal T}_2$ and not in ${\cal T}_1$.

@@

\begin{figure*}[tb!]
  \centering
  \includegraphics[width=\textwidth]{figures/diff}
  \caption{\framework comparison and accuracy results view.}
  \label{f:diff}
\end{figure*}

### 2018-03-14 tex/merf/OldSubmission/abstract.tex
old submission files

Rule-based techniques and tools to extract {\em entities} and {\em relational entities} from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.  These techniques and tools require expertise in linguistics and programming and lack support of Arabic {\em morphological analysis} which is key to process Arabic text.
In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.  \framework provides a friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines {\em tag types} and interactively associates them with regular expressions defined over Boolean formulae. Boolean formulae range over matches of Arabic morphological features, and synonymity features. Users define relations with tuples of subexpression matches and can associate code actions with subexpressions. \framework computes feature matches, regular expression matches, and constructs entities and relational entities from the user-defined relations. We evaluated our work with several case studies and compared with existing application-specific techniques. The results show that \framework requires shorter development time and effort compared to existing techniques and produces reasonably accurate results within a reasonable overhead in run time.

>>>>

Rule-based techniques and tools to extract {\em entities}  and {\em relational entities} from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.  These techniques and tools require expertise in linguistics and programming and lack support of Arabic {\em morphological analysis} which is key to process Arabic text.

In this work, we present MERF; a morphology-based entity and relational entity extraction framework for Arabic text.  MERF provides a user-friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines {\em tag types } and interactively associates them with regular expressions defined over Boolean formulae.  Boolean formulae range over matches of Arabic morphological features, and synonymity features.  Users define user defined relations with tuples of subexpression matches and can associate code actions with subexpressions.  MERF computes feature matches, regular expression matches, and constructs entities and relational entities from user defined relations.  We evaluated our work with several case studies and compared with existing application-specific techniques.  The results show that MERF requires shorter development time and effort compared to existing techniques and produces reasonably accurate results within a reasonable overhead in run time.

### 2018-03-14 tex/merf/OldSubmission/directionFigure.tex
old submission files

\transtrue
\setcode{utf8}
\setarab
\transfalse
\begin{figure}[tb]
\begin{center}
\relsize{-0.5}
%\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{\columnwidth}{!}{
    \begin{tabular}{|C{9.5cm}C{8cm}|} 
      \hline
  \begin{tabular}{R{9.5cm}}
  \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من} 
  \\ ~ \\
	$\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك} \RL{تقود} 
  \\ ~ \\
  $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا} $\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي} 
  \\
  \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  
  \\
  \multicolumn{1}{L{9.5cm}}{
   \arabfalse \transtrue
   \RL{من المستحيل ألا تلاحظ برج خليفة بالقرب من التقاطع الأول وأنت تقود سيارتك في شارع الشيخ زايد، حتّى وإن كانت هذه المرّة الأولى التي تسلك فيها هذا الطريق؛ يقع دبيّ مول على مقربة من هذا المبنى الذي يُعد الأطول في العالم.}
   \arabtrue \transfalse} 
  \\
  \multicolumn{1}{L{9.5cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.} 
  \end{tabular}
&
\resizebox{0.6\columnwidth}{!}{
  \input{figures/ergraphEx.tex}}
 \\ \hline
%\multicolumn{1}{c}{\multirow{1}{*}{\textbf{(a) Text with directions}}} &
%\multicolumn{1}{c}{\textbf{(b) Formula, matches, and entity-relation graph}} \\
    \end{tabular}
}
\end{center}
  \caption{Direction example with Arabic text, annotated with entities, transliteration, translation, and extracted relational entities in a graph} 
  \label{fig:intromotiv}
\end{figure}
\transtrue
\setcode{standard}

>>>>

\transtrue
\setcode{utf8}
\setarab
\transfalse
\begin{figure*}[tb]
%  \begin{center}
%\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{2\columnwidth}{!}{
    \begin{tabular}{|C{10cm}C{10cm}|} 
      \hline
  \begin{tabular}{R{10cm}}
  \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من} 
  \\ ~ \\
	$\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك} \RL{تقود} 
  \\ ~ \\
  $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا} $\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي} 
  \\
  \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  
  \\
  \multicolumn{1}{L{10cm}}{
   \arabfalse \transtrue
   \RL{من المستحيل ألا تلاحظ برج خليفة بالقرب من التقاطع الأول وأنت تقود سيارتك في شارع الشيخ زايد، حتّى وإن كانت هذه المرّة الأولى التي تسلك فيها هذا الطريق؛ يقع دبيّ مول على مقربة من هذا المبنى الذي يُعد الأطول في العالم.}
   \arabtrue \transfalse} 
  \\
  \multicolumn{1}{L{10cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.} 
  \end{tabular}
&
\resizebox{0.9\columnwidth}{!}{
  \input{figures/ergraphEx.tex}}
 \\ \hline
%\multicolumn{1}{c}{\multirow{1}{*}{\textbf{(a) Text with directions}}} &
%\multicolumn{1}{c}{\textbf{(b) Formula, matches, and entity-relation graph}} \\
    \end{tabular}
}
%  \end{center}
  \caption{Direction example with Arabic text, annotated with entities, transliteration, translation, and extracted relational entities in a graph} 
  \label{fig:intromotiv}
\end{figure*}
\transtrue
\setcode{standard}

### 2018-03-14 tex/merf/OldSubmission/discussion.tex
old submission files

The results show that \framework 
provides a friendly environment to develop entity and relational
entity extraction tasks with acceptable 
accuracy and runtime overheads compared to task specific applications. 
%
\framework requires the user to understand and interact with 
basic linguistic concepts such as readable values of morphological 
features, sequences, repetitions, and bounded repetitions. 
The user interacts with the MBF editor to specify basic concepts
and visualize their matches over highlighted text. 
%
Then, the user interacts with the MRE editor to specify 
sequences of the concepts and visualize the matches
in a graph, in conjunction with the highlighted text.

@@

%\subsection{ Threats to validity}
%As with any GUI tool, it takes time for the user to get acquainted with the 
%user friendly interface. 
%Relating the subexpressions to their suggested names by \framework might 
%not be directly intuitive to the user. 
%We address the above two concerns with providing the users with 
%a short tutorial video on how to use the tool.

>>>>

The results show that \framework 
provides a friendly environment to develop entity and relational
entity extraction tasks with acceptable 
accuracy and runtime overheads compared to task specific applications. 
%
\framework requires the user to understand and interact with 
basic linguistic concepts such as readable values of morphological 
features, sequences, repetitions, and bounded repetitions.
The user interacts with the MBF editor to specify basic concepts
and visualize their matches over highlighted text. 
%
Then the user interacts with the MRE editor to specify 
sequences of the concepts and visualize the matches
in a graph form, in conjunction with the highlighted text.

@@

\subsection{ Threats to validity}
As with any GUI tool, it takes time for the user to get acquainted with the 
user friendly interface. 
Relating the subexpressions to their suggested names by \framework might 
not be directly intuitive to the user. 
We address the above two concerns with providing the users with 
a short tutorial video on how to use the tool.
